dispRity manual

### 2017-01-10 doc/dispRity-manual.Rmd
Manual update for 0.3.0

<!-- TODO: change to html link -->

@@

`time.series`
-------------

@@

The `dispRity` algorithm decomposes the metrics functions into three levels that correspond to the dimensions of the output of each metric function.
For more details on this algorithm please refer to the [metric vignette](https://github.com/TGuillerme/dispRity).

<!-- TODO: change to html link -->

@@

@@

```{r, fig.width=6, fig.height=6}
## Graphical parameters
op <- par(mfrow = c(2,2), bty = "n")
## Plotting continuous disparity results
plot(disp_time_binsBS, type = "continuous")
## Plotting discrete disparity results
plot(disp_crown_stemBS, type = "box")
## Plotting the disparity using lines for the rarefaction level of 20 elements only
plot(disp_crown_stemBS, type = "lines", rarefaction = 20)
## Plotting the disparity using polygons while also displaying the number of elements
plot(disp_crown_stemBS, type = "polygon", elements = TRUE)
```

@@

```{r, fig.width=6, fig.height=6}
## Graphical options
par(bty = "n")
## Plotting the continuous disparity with a fixed y axis
plot(disp_time_slicesBS, ylim = c(0.5,2))
## Adding the discrete data using the same fixed y axis to have the results at the same scale
plot(disp_time_binsBS, discrete_type = "line", ylim=c(0.5,2), xlab="", ylab="", add = TRUE)
```

<!-- TODO: fix the placement bug with add = TRUE! -->

@@

```{r, fig.width=6, fig.height=6}
## Graphical options
par(bty = "n")
## Plotting the rarefaction curves
plot(disp_crown_stemBS, rarefaction = TRUE)
```
<!--

@@

Finally, the `dispRity` package allows to apply some tests to the data in order to properly test hypothesis.
The function `test.dispRity` works in a similar way as the `dispRity` function: it intakes a `dispRity` object (containing disparity measurements), a `test` and a `comparisons` argument.
The `test` argument can be any statistical or non-statistical test to apply to the disparity object. It can be common test function (e.g. `stats::t.test`) or some functions implemented in `dispRity` (see `dispRity.test`) or even user defined functions. The `comparisons` argument must indicate the way the test should be applied to the data:

@@

@@

```{r}
## Performing a pairwise t-test to the test a difference in disparity between
## crown and stem mammals (note that "pairwise" is default)
test.dispRity(disp_crown_stemBS, test = t.test)

## The same test but with the detailed output
test.dispRity(disp_crown_stemBS, test = t.test, details = TRUE)

## A wilcoxon rank test applied to sliced disparity the data sequentially with
## the Bonferroni correction.
test.dispRity(disp_time_slicesBS, test = wilcox.test,
    comparisons = "sequential", correction = "bonferroni")

## An analysis of variance (aov) applied to the same data
test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all")

## Note that in this case, the output is the regular aov output:
slice_aov <- test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all")
class(slice_aov) ; summary(slice_aov)

## Measuring the overlap between distributions in the time bins (using the
## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff)
test.dispRity(disp_time_binsBS, test = bhatt.coeff)

```

utilities
---------

### `dispRity` object utilities  <a name="dispRity.utilities"></a>

export(make.dispRity)
export(fill.dispRity)
export(matrix.dispRity)
export(get.series.dispRity)
export(extract.dispRity)
export(scale.dispRity)
export(sort.dispRity)

@@

## `tree.age`

@@

@@

Some of the core functions from the package have a `parallel` option.
 -->

>>>>

@@

`time.series`  <a name="time.series"></a>
-------------

@@

The `dispRity` algorithm decomposes the metrics functions into three levels that correspond to the dimensions of the output of each metric function.
For more details on this algorithm please refer to the [metric vignette](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-metric.Rmd).

@@

<a name="disparity.calculation"></a>

@@

```{r, fig.width=6, fig.height=6}
## Graphical parameters
op <- par(mfrow = c(2,2), bty = "n")
## Plotting continuous disparity results
plot(disp_time_binsBS, type = "continuous")
## Plotting discrete disparity results
plot(disp_crown_stemBS, type = "box")
## Plotting the disparity using lines for the rarefaction level of 20 elements only
plot(disp_crown_stemBS, type = "line", rarefaction = 20)
## Plotting the disparity using polygons while also displaying the number of elements
plot(disp_crown_stemBS, type = "polygon", elements = TRUE)
```

@@

```{r, fig.width=6, fig.height=6}
## Graphical options
par(bty = "n")
## Plotting the continuous disparity with a fixed y axis
plot(disp_time_slicesBS, ylim = c(0.5,2))
## Adding the discrete data using the same fixed y axis to have the results at the same scale
plot(disp_time_binsBS, type = "line", ylim=c(0.5,2), xlab="", ylab="", add = TRUE)
```

@@

```{r, fig.width=6, fig.height=6}
## Graphical options
par(bty = "n")
## Plotting the rarefaction curves
plot(disp_crown_stemBS, rarefaction = TRUE)
```

@@

Finally, the `dispRity` package allows to apply some test functions to the calculated disparity in order to test hypothesis.
The function `test.dispRity` works in a similar way as the `dispRity` function: it intakes a `dispRity` object, a `test` and a `comparisons` argument.

The `comparisons` argument must indicate the way the test should be applied to the data:

@@

The `test` argument can be any statistical or non-statistical test to apply to the disparity object.
It can be common statistical test functions (e.g. `stats::t.test`), some functions implemented in `dispRity` (e.g. see `?null.test`) or any function defined by the user.

@@

```{r}
## Performing a t-test to the test a difference in disparity between crown and stem mammals
test.dispRity(disp_crown_stemBS, test = t.test)

## Performing the same test but with the detailed t.test output
test.dispRity(disp_crown_stemBS, test = t.test, details = TRUE)

## Performing a Wilcoxon test applied to sliced disparity the data sequentially with correction
test.dispRity(disp_time_slicesBS, test = wilcox.test, comparisons = "sequential",
    correction = "bonferroni")

## Measuring the overlap between distributions in the time bins (using the implemented
## Bhattacharyya Coefficient function - see ?bhatt.coeff)
test.dispRity(disp_time_binsBS, test = bhatt.coeff)
```

It is also possible to apply some more *complex* tests that have their own output class (like `stats::aov`).
The results can then be analysed as usual using the associated `summary` S3 method:

```{r}
## Performing and ANOVA applied to the same data
slice_aov <- (test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all"))

## The output is a regular ANOVA output...
class(slice_aov)

## That can be summarised using summary
summary(slice_aov)
```

utilities
---------
The package also provides some utilities functions to facilitate multidimensional analysis.

### `dispRity` object utilities  <a name="dispRity.utilities"></a>
The first series of utilities are functions for manipulating `dispRity` objects:

#### `make.dispRity`
This function creates empty `dispRity` objects.

```{r}
## Creating an empty dispRity object
make.dispRity()
## Creating an "empty" dispRity object with a matrix
(disparity_obj <- make.dispRity(matrix(rnorm(20), 5, 4)))
```

####  `fill.dispRity`
This function initialises a `dispRity` object and generating it's call properties.

```{r}
## The dispRity object's call is indeed empty
disparity_obj$call
## Creating an empty dispRity object
(disparity_obj <- fill.dispRity(disparity_obj))
## The dipRity object has now the proper minimal attributes
disparity_obj$call
```

####  `matrix.dispRity`
This function extracts a specific matrix from a disparity object. The matrix can be one of the bootstrapped or/and rarefied one.

```{r}
## Extracting the matrix containing the coordinates for the crown mammals
str(matrix.dispRity(disp_crown_stemBS, "Group.crown"))
## Extracting the 3rd boostrapped matrix with the 10th rarefaction level from the same group
str(matrix.dispRity(disp_crown_stemBS, series = 1, bootstrap = 3, rarefaction = 10))
```

#### `get.series.dispRity`
This function creates a dispRity object that contains only elements from one specific series.

```{r}
## Extracting all the data for the crown mammals
(crown_mammals <- get.series.dispRity(disp_crown_stemBS, "Group.crown"))
## The object keeps the properties of the parent object but is composed of only one series
length(crown_mammals$series)
```

#### `extract.dispRity`
This function extracts the calculated disparity values of a specific matrix.

```{r}
## Extracting the observed disparity (default)
extract.dispRity(disp_crown_stemBS)

## Extracting the disparity from the bootstrapped values from the 10th rarefaction level from the
## second series (stem mammals)
extract.dispRity(disp_crown_stemBS, observed = FALSE, series = 2, rarefaction = 10)
```

#### `scale.dispRity`
This is the S3 method of `scale` (scaling and/or centring) that can be applied to the disparity data of a `dispRity` object.

```{r}
## Getting the disparity values of the time series
summary(disp_time_slicesBS)

## Scaling the same disparity values
summary(scale(disp_time_slicesBS, scale = TRUE))
```

#### `sort.dispRity`
This is the S3 method of `sort` for sorting the series alphabetically (default) or following a specific pattern.

```{r}
## Sorting the disparity series in inverse alphabetic order
summary(sort(disp_time_slicesBS, decreasing = TRUE))
```

@@

#### `tree.age`

@@

#### `slice.tree`

This function is at the core of `time.series` when using the time slicing method.
It allows to slice a phylogenetic tree at a certain time point and selecting the nodes or tips following the models detailed in the `time.series` [section above](#time.series).

```{r, fig.width=8, fig.height=6}
set.seed(1)
## Generating a random ultrametric tree
tree <- rcoal(20)
## Adding some node labels
tree$node.label <- letters[1:19]
## Adding a root age (as the maximum tree age)
tree$root.time <- max(tree.age(tree)$ages)

## Creating a deltran tree at the slice 0.25
tree_25 <- slice.tree(tree, age = 0.25, "deltran")
## Creating a deltran tree at the slice 0.50
tree_50 <- slice.tree(tree, age = 0.50, "acctran")

## Illustrating the results
par(mfrow = c(1,3))
plot(tree, main = "Full tree", cex = 1.5)
nodelabels(tree$node.label, cex = 1.1)
axisPhylo()
plot(tree_25, main = "Slice at time 0.25", cex = 1.5)
axisPhylo()
plot(tree_50, main = "Slice at time 0.50", cex = 1.5)
axisPhylo()
```

#### `sim.morpho`
This function (and it's associated utilities) allows to generate realistic discrete morphological matrices.
See the [associated vignette](https://github.com/TGuillerme/dispRity/blob/master/doc/dispRity-simulate_data.Rmd) for more details.

Other utilities functions are also available within the package and more will be added in the future. For the full list, refer to the package's functions list within `R`.

@@

`dispRity` and `boot.matrix` have also `parallel` option.
This option uses the package `snow` to run the `lapply` loops within the package in parallel.
In both functions, this option is ignored when missing, otherwise, it takes as option, the number of threads (cores) and the virtual connection process (VCP) name following the function `makeCluster`.

For example, we can speed up the [disparity calculations of the rarefied data](#disparity.calculation) by using 4 different threads and a `"SOCK"` connection (although `"MPI"` is much more recommended but needs more specific installation).

```{r, eval=FALSE}
dispRity(crown_stemBS, metric = c(median, centroids), parallel = c(4, "SOCK"))
```
